---
title: |
  | _Aedes aegypti_ mRNA expression --
  | Differential expression downstream analyses
pagetitle: "DE"
author: "Jelmer Poelstra (poelstra.1@osu.edu), MCIC Wooster, OSU"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: cerulean
    highlight: kate
    toc: true
    toc_float: true
    fig_caption: true
    anchor_sections: true
    df_print: kable
    css: html_page.css
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  eval = TRUE,
  cache = FALSE,
  warning = FALSE,
  message = FALSE,
  out.width = "80%"
)
knitr::opts_knit$set(root.dir = here::here())
```

```{r}
## Install/load packages
if(! "pacman" %in% installed.packages()) install.packages("pacman")
packages <- c("tidyverse",        # Misc. data manipulation and plotting
              "here",             # Managing file paths
              "VennDiagram",      # Venn Diagram
              "pheatmap",         # Heatmaps
              "RColorBrewer",     # Plotting colors
              "plotly",           # Interactive plots
              "glue",             # Pasting text
              "DT", "kableExtra") # Tables             
pacman::p_load(char = packages)
```

```{r}
## Source scripts with functions
source(here("scripts/report_funs.R"))
```

```{r}
## Set contrasts of interest
contrasts <- c("lvp06_vir06", "lvp24_vir24")
contrasts_all <- c("akh06_vir06", "akh24_vir24",
                   "lvp06_vir06", "lvp24_vir24",
                   "lvp06_akh06", "lvp24_akh24")
contrast_labs <- c("6 hpm", "24 hpm")

## Define tissues
tissues <- c("ab", "ht", "lrt")
tissue_labs <- c("Ab", "HT", "LRT")
```

```{r}
## Set colors for Venn diagram
contrast_cols <- brewer.pal(4, "Pastel2")[1:2]
tissue_cols <- brewer.pal(3, "Set2")
```

```{r}
## Define input files
indir_DE <- here("results", "DE")
indir_counts <- here("results", "counts")

gene_file <- here(indir_DE, "gene_info.txt")
DE_file <- here(indir_DE, "DE_all.txt")
counts_file <- here(indir_counts, "counts_all.txt")
```

```{r}
## Read input files
gene_df <- read_tsv(gene_file, show_col_types = FALSE) %>%
  select(gene_id, description) %>%
  mutate(description = str_trunc(description, width = 50))

DE_res_all <- read_tsv(DE_file, show_col_types = FALSE) %>%
  mutate(tissue = gsub(".*\\d(\\w+)", "\\1", treat_a),
         tissue = ifelse(sugar == "sug12", "lrt12", tissue),
         trt_a = gsub("(.*\\d+)\\w+", "\\1", treat_a),
         trt_b = gsub("(.*\\d+)\\w+", "\\1", treat_b),
         contrast = paste0(trt_a, "_", trt_b),
         contrast_full = paste0(contrast, "_", tissue),
         sig = ifelse(padj < 0.05, TRUE, FALSE)) %>%
  filter(contrast %in% contrasts) %>%            # Only select focal contrasts!
  select(gene_id, lfc, pvalue, padj, sig,
         trt_a, trt_b, tissue, sugar, contrast, contrast_full) %>%
  left_join(gene_df, by = "gene_id")
DE_res <- DE_res_all %>% filter(tissue %in% tissues)

count_df_all <- read_tsv(counts_file, show_col_types = FALSE)
count_df <- count_df_all %>% filter(sugar != "sug12", !grepl("akh", treatment))
```

```{r}
## Sample metadata
meta_df <- count_df %>%
  distinct(sample, .keep_all = TRUE) %>%
  select(sample, tissue, sugar, treatment) %>%
  as.data.frame()
rownames(meta_df) <- meta_df$sample
```

```{r}
## Make a df with all pairwise contrasts
contrast_df <- DE_res %>%
  mutate(time_a = gsub(".*(\\d\\d)", "\\1", trt_a),
         time_b = gsub(".*(\\d\\d)", "\\1", trt_b),
         time_contrast = factor(paste0(time_a, "_", time_b),
                                levels = c("06_06", "24_24"))) %>%
  distinct(contrast_full, .keep_all = TRUE) %>% 
  select(contrast_full, contrast, trt_a, trt_b, tissue, sugar, time_contrast)
```

<br>

I'm focusing on the following pairwise contrasts:

- LVP vs. virgin at 6 hpm
- LVP vs. virgin at 24 hpm

Abbreviations:

- `DEGs`: Differentially expressed genes
- `padj`: Multiple-testing corrected (adjusted) p-value
- `lfc`: Log2-fold change in mean expression level

**Note: heatmaps show log-10 transformed values.**

<br>

----

## DEGs per contrast

### Number of DEGs per contrast

```{r}
sig_by_comp <- DE_res %>%
  mutate(contrast_full = factor(contrast_full), sig = factor(sig)) %>% 
  group_by(contrast_full, sig, .drop = FALSE) %>%
  count(name = "nsig") %>%
  filter(sig == TRUE) %>% 
  ungroup() %>%
  select(contrast_full, nsig) %>%
  full_join(contrast_df, by = "contrast_full") %>% 
  select(trt_a, trt_b, tissue, time_contrast, nsig) %>%
  pivot_wider(id_cols = c(trt_a, trt_b, time_contrast),
              names_from = tissue,
              values_from = nsig) %>%
  arrange(time_contrast, trt_a, trt_b) %>%
  select(-time_contrast)
```

```{r}
sig_by_comp <- DE_res %>%
  mutate(contrast_full = factor(contrast_full),
         sig = factor(sig),
         direction = ifelse(lfc < 0, "down", "up")) %>% 
  group_by(contrast_full, sig, direction, .drop = FALSE) %>%
  count(name = "nsig") %>%
  filter(sig == TRUE) %>% 
  ungroup() %>%
  select(contrast_full, direction, nsig) %>%
  full_join(contrast_df, by = "contrast_full") %>% 
  select(trt_a, trt_b, tissue, time_contrast, direction, nsig) %>%
  pivot_wider(id_cols = c(trt_a, trt_b, time_contrast, direction),
              names_from = tissue,
              values_from = nsig,
              values_fill = 0) %>%
  arrange(time_contrast, trt_a, trt_b) %>%
  select(-time_contrast) %>%
  filter(!is.na(direction))
```

```{r}
## Barplot
DE_res %>%
  filter(sig == TRUE, contrast %in% contrasts[1:4]) %>%
  ggplot(aes(x = contrast, fill = tissue)) +
  geom_bar(position = position_dodge(preserve = "single"),
           color = "grey40") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  scale_x_discrete(labels = contrast_labs, drop = FALSE) +
  scale_fill_brewer(palette = "Set2",
                    labels = tissue_labs,
                    name = "Tissue",
                    drop = FALSE) +
  labs(x = "", y = "Number of DEGs") +
  theme_classic() +
  theme(legend.position = "top")
```

<br>

In the table below, "down" means lower expression in LVP than in VIR,
and "up" means higher expression in LVP than in VIR:

```{r}
## Table
sig_by_comp %>% 
  rename("group A" = trt_a, "group B" = trt_b,
         abdomen = ab, "head & thorax" = ht) %>%
  make_kable() 
```

### Volcano plots {.tabset .tabset-fade .tabset-pills}

- Only showing significantly differentially expressed genes.
- A higher value on the y-axis is more highly significantly DE.
- A higher Log-fold change (LFC) is a bigger difference between the mean expression
  levels of the group, with LFC > 0 meaning that the first-mentioned group in
  the contrast has a higher expression level than the second.
- **The plots are interactive so you see info about each point (e.g. gene ID)
  by hovering over it. You can also exclude contrasts by clicking on them in
  the legend on the right.**

```{r}
volc_plot <- function(DE_df, ftissue) {
  p <- DE_df %>%
    filter(sig == TRUE, tissue == ftissue) %>%
    ggplot(aes(x = lfc, y = -log10(pvalue), color = contrast,
               text = glue("Gene: {gene_id}
                           Description: {description}
                           Contrast: {contrast}"))) +
    geom_point() +
    geom_vline(xintercept = 0, color = "grey30") +
    scale_color_brewer(palette = "Dark2") +
    labs(x = "Log-fold change (>0: 1st treatment has higher counts)") +
    theme_bw(base_size = 14)
  ggplotly(p, tooltip = "text")
}
```

#### Abdomen

```{r, out.width="90%"}
volc_plot(DE_res, ftissue = "ab")
```

<br>

-----

#### Head & thorax

```{r, out.width="90%"}
volc_plot(DE_res, ftissue = "ht")
```

<br>

-----

#### LRT (3%)

```{r, out.width="90%"}
volc_plot(DE_res, ftissue = "lrt")
```

<br>

-----

## Top 10 DEG lists

```{r}
## Create a df with top-10 DE genes
top10 <- function(DE_df, ftissue, fcontrast,
                  contrast_col = TRUE, tissue_col = FALSE) {
  DE_df <- DE_df %>%
    filter(tissue %in% ftissue, contrast %in% fcontrast, padj < 0.05) %>%
    arrange(padj) %>%
    select(-pvalue, -sig, -sugar, -contrast_full, -trt_a, -trt_b) %>% 
    slice(1:10) %>%
    mutate(padj = ifelse(padj < 0.001, f_sci(padj), f_dec(padj)))
  
  if (contrast_col == FALSE) DE_df <- DE_df %>% select(-contrast)
  if (tissue_col == FALSE) DE_df <- DE_df %>% select(-tissue)
  
  return(DE_df)
}
```

```{r}
## Heatmap plot wrapper
plot_heat2 <- function(IDs, take_log = TRUE) {
  
  IDs <- unique(IDs)
  
  fcount_mat <- count_df %>%
    filter(gene_id %in% IDs) %>%
    select(gene_id, sample, count) %>%
    pivot_wider(names_from = sample, values_from = count) %>%
    column_to_rownames("gene_id") %>%
    as.matrix()

  if (take_log == TRUE) {
    fcount_mat <- log10(fcount_mat)
    fcount_mat[fcount_mat == -Inf] <- 0
  }
  
  plot_heatmap(IDs, fcount_mat, meta_df, groups = c("tissue", "treatment"))
}
```

```{r}
## Single-gene boxplot
plot_box <- function(ID) {
  
  descrip <- gene_df %>%
    filter(gene_id == ID) %>%
    pull(description) %>%
    str_trunc(width = 50)
  
  p <- count_df %>%
    filter(gene_id %in% ID) %>%
    ggplot() +
    aes(y = count, x = treatment, color = treatment) +
    geom_boxplot(outlier.shape = NA) +
    geom_point(position = position_jitter(w = 0.15, h = 0),
               size = 1.5) +
    facet_wrap(vars(tissue)) +
    scale_y_continuous(labels = scales::comma,
                       expand = expansion(mult = c(0.001, 0.2))) +
    scale_color_brewer(palette = "Paired") +
    labs(title = ID, subtitle = descrip,
         x = NULL, y = "Normalized count") +
    guides(color = "none") +
    theme_bw(base_size = 13) +
    theme(legend.position = "right",
          plot.title = element_text(hjust = 0.5, face = "bold.italic"),
          plot.subtitle = element_text(hjust = 0.5, face = "plain"),
          panel.grid.major.x = element_blank(),
          panel.grid.minor.y = element_blank(),
          plot.margin = unit(c(0.5, 0.5, 1, 0.5), "cm"))
  
  print(p)
}
```

### Across all tissues & contrasts {.tabset .tabset-fade .tabset-pills}

```{r}
t10 <- top10(DE_res, tissues, contrasts, tissue_col = TRUE)
```

#### Table

<br>

```{r}
make_dt(t10)
```

#### Heatmap

```{r}
if (nrow(t10) > 1) plot_heat2(t10$gene_id)
```

#### Boxplots

```{r}
walk(unique(t10$gene_id), plot_box)
```

### Across all tissues by contrast {.tabset .tabset-fade .tabset-pills}

<br>

#### `r contrasts[1]` {.tabset .tabset-fade .tabset-pills}

```{r}
t10 <- top10(DE_res, tissues, contrasts[1],
             contrast_col = FALSE, tissue_col = TRUE)
```

##### Table

<br>

```{r}
make_dt(t10)
```

##### Heatmap

```{r}
if (nrow(t10) > 1) plot_heat2(t10$gene_id)
```

##### Boxplots

```{r}
walk(unique(t10$gene_id), plot_box)
```

#### `r contrasts[2]` {.tabset .tabset-fade .tabset-pills}

```{r}
t10 <- top10(DE_res, tissues, contrasts[2],
             contrast_col = FALSE, tissue_col = TRUE)
```

##### Table

<br>

```{r}
make_dt(t10)
```

##### Heatmap

```{r}
if (nrow(t10) > 1) plot_heat2(t10$gene_id)
```

##### Boxplots

```{r}
walk(unique(t10$gene_id), plot_box)
```

### Across all contrasts by tissue {.tabset .tabset-fade .tabset-pills}

<br>

#### `r tissues[1]` {.tabset .tabset-fade .tabset-pills}

```{r}
t10 <- top10(DE_res, tissues[1], contrasts)
```

##### Table

<br>

```{r}
make_dt(t10)
```

##### Heatmap

```{r}
if (nrow(t10) > 1) plot_heat2(t10$gene_id)
```

##### Boxplots

```{r}
walk(unique(t10$gene_id), plot_box)
```

#### `r tissues[2]` {.tabset .tabset-fade .tabset-pills}

```{r}
t10 <- top10(DE_res, tissues[2], contrasts)
```

##### Table

<br>

```{r}
make_dt(t10)
```

##### Heatmap

```{r}
if (nrow(t10) > 1) plot_heat2(t10$gene_id)
```

##### Boxplots

```{r}
walk(unique(t10$gene_id), plot_box)
```

#### `r tissues[3]` {.tabset .tabset-fade .tabset-pills}

```{r}
t10 <- top10(DE_res, tissues[3], contrasts)
```

##### Table

<br>

```{r}
make_dt(t10)
```

##### Heatmap

```{r}
if (nrow(t10) > 1) plot_heat2(t10$gene_id)
```

##### Boxplots

```{r}
walk(unique(t10$gene_id), plot_box)
```

### Abdomen {.tabset .tabset-fade .tabset-pills}

<br>

```{r}
tissue <- "ab"
```

#### `r contrasts[1]` {.tabset .tabset-fade .tabset-pills}

```{r}
t10 <- top10(DE_res, tissue, contrasts[1])
```

##### Table

<br>

```{r}
make_dt(t10)
```

##### Heatmap

```{r}
if (nrow(t10) > 1) plot_heat2(t10$gene_id)
```

##### Boxplots

```{r}
walk(unique(t10$gene_id), plot_box)
```

#### `r contrasts[2]` {.tabset .tabset-fade .tabset-pills}

```{r}
t10 <- top10(DE_res, tissue, contrasts[2])
```

##### Table

<br>

```{r}
make_dt(t10)
```

##### Heatmap

```{r}
if (nrow(t10) > 1) plot_heat2(t10$gene_id)
```

##### Boxplots

```{r}
walk(unique(t10$gene_id), plot_box)
```

### Head & thorax {.tabset .tabset-fade .tabset-pills}

```{r}
tissue <- "ht"
```

<br>

#### `r contrasts[1]` {.tabset .tabset-fade .tabset-pills}

```{r}
t10 <- top10(DE_res, tissue, contrasts[1])
```

##### Table

<br>

```{r}
make_dt(t10)
```

##### Heatmap

```{r}
if (nrow(t10) > 1) plot_heat2(t10$gene_id)
```

##### Boxplots

```{r}
walk(unique(t10$gene_id), plot_box)
```

#### `r contrasts[2]` {.tabset .tabset-fade .tabset-pills}

```{r}
t10 <- top10(DE_res, tissue, contrasts[2])
```

##### Table

<br>

```{r}
make_dt(t10)
```

##### Heatmap

```{r}
if (nrow(t10) > 1) plot_heat2(t10$gene_id)
```

##### Boxplots

```{r}
walk(unique(t10$gene_id), plot_box)
```

### LRT (3%) {.tabset .tabset-fade .tabset-pills}

<br>

```{r}
tissue <- "lrt"
```

#### `r contrasts[1]` {.tabset .tabset-fade .tabset-pills}

```{r}
t10 <- top10(DE_res, tissue, contrasts[1])
```

##### Table

<br>

```{r}
make_dt(t10)
```

##### Heatmap

```{r}
if (nrow(t10) > 1) plot_heat2(t10$gene_id)
```

##### Boxplots

```{r}
walk(unique(t10$gene_id), plot_box)
```

#### `r contrasts[2]` {.tabset .tabset-fade .tabset-pills}

```{r}
t10 <- top10(DE_res, tissue, contrasts[2])
```

##### Table

<br>

```{r}
make_dt(t10)
```

##### Heatmap

```{r}
if (nrow(t10) > 1) plot_heat2(t10$gene_id)
```

##### Boxplots

```{r}
walk(unique(t10$gene_id), plot_box)
```

## DEGs across multiple contrasts

### Any contrast combination {.tabset .tabset-fade .tabset-pills}

- `sign. in n. contrasts` is the number of contrasts in which the gene is DE.
- Values in columns with contrast names are adjusted p-values.

```{r}
## Function to create a df with genes significant in multiple focal contrasts
multisig <- function(DE_df, tissues, minsig = 2) {
  
  if (length(tissues) > 1) DE_df <- DE_df %>% mutate(contrast = contrast_full) 
  
  DE_df %>% 
    filter(sig == TRUE, tissue %in% tissues) %>%
    add_count(gene_id, name = "n_sig") %>%
    filter(n_sig >= minsig) %>% 
    select(gene_id, n_sig, contrast, padj) %>%
    pivot_wider(id_cols = c(gene_id, n_sig),
                names_from = contrast, values_from = padj) %>%
    arrange(desc(n_sig), gene_id) %>%
    mutate(n_sig = as.integer(n_sig)) %>% 
    rename("sign. in n contrasts" = n_sig) %>%
    left_join(gene_df, by = "gene_id")
}
```

<br>

#### Abdomen {.tabset .tabset-fade .tabset-pills}

<br>

```{r}
msig <- multisig(DE_res, tissues = "ab")
```

##### Table

```{r}
make_dt(msig, numr_cols = colnames(msig)[3:(ncol(msig) - 1)])
```

##### Heatmap for top-100

```{r}
if (nrow(msig) > 1) plot_heat2(msig$gene_id[1:min(100, length(msig$gene_id))])
```

##### Boxplots for top-10

```{r}
walk(msig$gene_id[1:min(10, length(msig$gene_id))], plot_box)
```

#### Head & thorax {.tabset .tabset-fade .tabset-pills}

<br>

```{r}
msig <- multisig(DE_res, tissues = "ht")
```

##### Table

```{r}
if(nrow(msig) > 0) make_dt(msig, numr_cols = colnames(msig)[3:(ncol(msig) - 1)])
```

##### Heatmap for top-100

```{r}
if(nrow(msig) > 1) plot_heat2(msig$gene_id[1:min(100, length(msig$gene_id))])
```

##### Boxplots for top-10

```{r}
if(nrow(msig) > 0) walk(msig$gene_id[1:min(10, length(msig$gene_id))], plot_box)
```

#### LRT (3%) {.tabset .tabset-fade .tabset-pills}

<br>

```{r}
msig <- multisig(DE_res, tissues = "lrt")
```

##### Table

```{r}
if(nrow(msig) > 0) make_dt(msig, numr_cols = colnames(msig)[3:(ncol(msig) - 1)])
```

##### Heatmap for top-100

```{r}
if(nrow(msig) > 1) plot_heat2(msig$gene_id[1:min(100, length(msig$gene_id))])
```

##### Boxplots for top-10

```{r}
if(nrow(msig) > 0) walk(msig$gene_id[1:min(10, length(msig$gene_id))], plot_box)
```

#### All tissues {.tabset .tabset-fade .tabset-pills}

<br>

```{r}
msig <- multisig(DE_res, tissues = tissues)
```

##### Table

```{r}
make_dt(msig, numr_cols = colnames(msig)[3:(ncol(msig) - 1)])
```

##### Heatmap for top-100

```{r}
plot_heat2(msig$gene_id[1:min(100, length(msig$gene_id))])
```

##### Boxplots for top-10

```{r}
walk(msig$gene_id[1:min(10, length(msig$gene_id))], plot_box)
```

### DE at both time points {.tabset .tabset-fade .tabset-pills}

```{r}
## Function to create a df with genes significant in multiple focal contrasts
multisig_time <- function(DE_df, ftissue) {
  msig <- DE_df %>%
    mutate(timepoint = sub("\\w+(\\d\\d)", "t\\1", trt_a),
           contrast = gsub("\\d", "", contrast)) %>% 
    filter(sig == TRUE, tissue == ftissue) %>%
    select(gene_id, lfc, padj, contrast, timepoint)
  
  if (length(unique(msig$timepoint)) > 1) {
    msig <- msig %>%  
      pivot_wider(id_cols = c(gene_id, contrast),
                  names_from = timepoint,
                  values_from = c(padj, lfc)) %>%
      mutate(padj_mean = mean(c(padj_t06, padj_t24), na.rm = TRUE)) %>% 
      filter(!is.na(padj_t06), !is.na(padj_t24)) %>%
      left_join(gene_df, by = "gene_id") %>%
      arrange(padj_mean) %>%
      select(-padj_mean)
    
    return(msig)
  
  } else {
    message("Only found significant results for 1 timepoint")
    return(data.frame())
  }
}
```

<br>

#### Abdomen {.tabset .tabset-fade .tabset-pills}

<br>

```{r}
bothtimes <- multisig_time(DE_res, ftissue = "ab")
```

##### Table

```{r}
bothtimes %>% make_dt(numr_cols = c("padj_t06", "padj_t24"))
```

##### Heatmap for top-100

```{r}
if (nrow(bothtimes) > 1)
    plot_heat2(bothtimes$gene_id[1:min(100, length(bothtimes$gene_id))])
```

##### Boxplots for top-10

```{r}
walk(bothtimes$gene_id[1:min(10, length(bothtimes$gene_id))], plot_box)
```

#### Head & thorax {.tabset .tabset-fade .tabset-pills}

<br>

```{r}
bothtimes <- multisig_time(DE_res, ftissue = "ht")
```

##### Table

```{r}
if (nrow(bothtimes) > 0) bothtimes %>% make_dt(numr_cols = c("padj_t06", "padj_t24"))
```

##### Heatmap for top-100

```{r}
if (nrow(bothtimes) > 1)
    plot_heat2(bothtimes$gene_id[1:min(100, length(bothtimes$gene_id))])
```

##### Boxplots for top-10

```{r}
if (nrow(bothtimes) > 0)
    walk(bothtimes$gene_id[1:min(10, length(bothtimes$gene_id))], plot_box)
```

#### LRT (3%) {.tabset .tabset-fade .tabset-pills}

<br>

```{r}
bothtimes <- multisig_time(DE_res, ftissue = "lrt")
```

##### Table

```{r}
bothtimes %>% make_dt(numr_cols = c("padj_t06", "padj_t24"))
```

##### Heatmap for top-100

```{r}
if (nrow(bothtimes) > 1)
    plot_heat2(bothtimes$gene_id[1:min(100, length(bothtimes$gene_id))])
```

##### Boxplots for top-10

```{r}
walk(bothtimes$gene_id[1:min(10, length(bothtimes$gene_id))], plot_box)
```

### DE in multiple tissues {.tabset .tabset-fade .tabset-pills}

<br>

```{r}
## Function to create a df with genes significant in multiple focal contrasts
multisig_tissue <- function(DE_df, fcontrast) {
  DE_df %>%
    filter(contrast == fcontrast, tissue != "lrt12") %>%
    add_count(gene_id, sig, name = "n_sig") %>%
    select(gene_id, n_sig, sig, lfc, padj, contrast, tissue) %>%
    pivot_wider(id_cols = c(gene_id, contrast, n_sig, sig),
                names_from = tissue,
                values_from = c(padj, lfc)) %>%
    filter(n_sig >= 2, sig == TRUE) %>% 
    left_join(gene_df, by = "gene_id") %>%
    select(-sig) %>%
    mutate(padj_mean = mean(c(padj_ab, padj_ht, padj_lrt), na.rm = TRUE),
           padj_ab = ifelse(padj_ab < 0.001, f_sci(padj_ab), f_dec(padj_ab)),
           padj_ht = ifelse(padj_ht < 0.001, f_sci(padj_ht), f_dec(padj_ht)),
           padj_lrt = ifelse(padj_lrt < 0.001, f_sci(padj_lrt), f_dec(padj_lrt))) %>%
    rename("sign. in n contrasts" = n_sig) %>%
    arrange(padj_mean) %>%
    select(-padj_mean, -contrast)
}
```

#### LVP vs Virgin -- 6 hpm {.tabset .tabset-fade .tabset-pills}

<br>

```{r}
bothtiss <- multisig_tissue(DE_res, fcontrast = "lvp06_vir06")
```

##### Table

```{r}
make_dt(bothtiss, numr_cols = c("padj_ab", "padj_ht", "padj_lrt"))
```

##### Heatmap for top-100

```{r}
if (nrow(bothtiss) > 1)
    plot_heat2(bothtiss$gene_id[1:min(100, length(bothtiss$gene_id))])
```

##### Boxplots for top-10

```{r}
if (nrow(bothtiss) > 0)
  walk(bothtiss$gene_id[1:min(10, length(bothtiss$gene_id))], plot_box)
```

#### LVP vs Virgin -- 24 hpm {.tabset .tabset-fade .tabset-pills}

<br>

```{r}
bothtiss <- multisig_tissue(DE_res, fcontrast = "lvp24_vir24")
```

##### Table

```{r}
make_dt(bothtiss, numr_cols = c("padj_ab", "padj_ht", "padj_lrt"))
```

##### Heatmap for top-100

```{r}
if (nrow(bothtiss) > 1)
    plot_heat2(bothtiss$gene_id[1:min(100, length(bothtiss$gene_id))])
```

##### Boxplots for top-10

```{r}
if (nrow(bothtiss) > 0)
  walk(bothtiss$gene_id[1:min(10, length(bothtiss$gene_id))], plot_box)
```

<br>

-----

## Venn diagrams

### DEG overlap across times {.tabset .tabset-fade .tabset-pills}

```{r}
## Function to get a vector with significantly DE genes for a pairwise contrast
get_genes <- function(fcontrast, ftissue, direction = "both",
                      DE_df) {
  DE_df <- DE_df %>% 
    filter(sig == TRUE, tissue == ftissue, contrast == fcontrast)
  if (direction == "up") DE_df <- DE_df %>% filter(lfc > 0)
  if (direction == "down") DE_df <- DE_df %>% filter(lfc < 0)
  
  DE_df %>% pull(gene_id)
}
```

#### Abdomen

```{r, results="hide"}
DE_genes <- map(.x = contrasts, .f = get_genes,
                ftissue = "ab", DE_df = DE_res)
names(DE_genes) <- contrast_labs

p <- venn.diagram(x = DE_genes, fill = contrast_cols, 
                  filename = NULL, output = TRUE, disable.logging = TRUE,
                  margin = 0.05, cat.cex = 0.9, cat.fontface = "bold")
grid::grid.draw(p)
```

#### Head & thorax

```{r, results="hide"}
DE_genes <- map(.x = contrasts, .f = get_genes,
                ftissue = "ht", DE_df = DE_res)
names(DE_genes) <- contrast_labs

p <- venn.diagram(x = DE_genes, fill = contrast_cols, 
                  filename = NULL, output = TRUE, disable.logging = TRUE,
                  margin = 0.05, cat.cex = 0.9, cat.fontface = "bold")
grid::grid.draw(p)
```

#### LRT (3%)

```{r, results="hide"}
DE_genes <- map(.x = contrasts, .f = get_genes,
                ftissue = "lrt", DE_df = DE_res)
names(DE_genes) <- contrast_labs

p <- venn.diagram(x = DE_genes, fill = contrast_cols, 
                  filename = NULL, output = TRUE, disable.logging = TRUE,
                  margin = 0.05, cat.cex = 0.9, cat.fontface = "bold")
grid::grid.draw(p)
```

### DEG overlap across tissues {.tabset .tabset-fade .tabset-pills}

#### 6 hpm - all

```{r, results="hide"}
DE_genes <- map(.x = tissues, .f = get_genes,
                fcontrast = "lvp06_vir06", DE_df = DE_res)
names(DE_genes) <- tissue_labs

p <- venn.diagram(x = DE_genes, fill = tissue_cols, 
                  filename = NULL, output = TRUE, disable.logging = TRUE,
                  margin = 0.05, cat.cex = 1.5, cat.fontface = "bold",
                  cat.pos = 0, cex = 1.5, fontface = "bold")
grid::grid.draw(p)
```

<br>

----

#### 6 hpm - up

```{r, results="hide"}
DE_genes <- map(.x = tissues, .f = get_genes, direction = "up",
                fcontrast = "lvp06_vir06", DE_df = DE_res)
names(DE_genes) <- tissue_labs

p <- venn.diagram(x = DE_genes, fill = tissue_cols, 
                  filename = NULL, output = TRUE, disable.logging = TRUE,
                  margin = 0.05, cat.cex = 1.5, cat.fontface = "bold",
                  cat.pos = 0, cex = 1.5, fontface = "bold")
grid::grid.draw(p)
```

<br>

----

#### 6 hpm - down

```{r, results="hide"}
DE_genes <- map(.x = tissues, .f = get_genes, direction = "down",
                fcontrast = "lvp06_vir06", DE_df = DE_res)
names(DE_genes) <- tissue_labs

p <- venn.diagram(x = DE_genes, fill = tissue_cols, 
                  filename = NULL, output = TRUE, disable.logging = TRUE,
                  margin = 0.05, cat.cex = 1.5, cat.fontface = "bold",
                  cat.pos = c(0, 0, 180), cex = 1.5, fontface = "bold")
grid::grid.draw(p)
```

<br>

----

#### 24 hpm - all

```{r, results="hide"}
DE_genes <- map(.x = tissues, .f = get_genes,
                fcontrast = "lvp24_vir24", DE_df = DE_res)
names(DE_genes) <- tissue_labs

p <- venn.diagram(x = DE_genes, fill = tissue_cols, 
                  filename = NULL, output = TRUE, disable.logging = TRUE,
                  margin = 0.05, cat.cex = 1.5, cat.fontface = "bold",
                  cex = 1.5, fontface = "bold")
grid::grid.draw(p)
```

<br>

----

#### 24 hpm - up

```{r, results="hide"}
DE_genes <- map(.x = tissues, .f = get_genes, direction = "up",
                fcontrast = "lvp24_vir24", DE_df = DE_res)
names(DE_genes) <- tissue_labs

p <- venn.diagram(x = DE_genes, fill = tissue_cols, 
                  filename = NULL, output = TRUE, disable.logging = TRUE,
                  margin = 0.05, cat.cex = 1.5, cat.fontface = "bold",
                  cat.pos = c(0, 0, 180), cex = 1.5, fontface = "bold")
grid::grid.draw(p)
```

<br>

----

#### 24 hpm - down

Lower expression levels in mated.

```{r, results="hide"}
DE_genes <- map(.x = tissues, .f = get_genes, direction = "down",
                fcontrast = "lvp24_vir24", DE_df = DE_res)
names(DE_genes) <- tissue_labs

p <- venn.diagram(x = DE_genes, fill = tissue_cols, 
                  filename = NULL, output = TRUE, disable.logging = TRUE,
                  margin = 0.05, cat.cex = 1.5, cat.fontface = "bold",
                  cat.pos = c(0, 0, 180), cex = 1.5, fontface = "bold")
grid::grid.draw(p)
```

<br>

----

## 3% vs. 12% sugar in LRT

### Barplot of DEGs

```{r}
DE_lrt <- DE_res_all %>%
  filter(tissue %in% c("lrt", "lrt12"),
         contrast %in% contrasts_all[1:4],
         sig == TRUE) %>%
  select(gene_id, contrast, padj, tissue) %>%
  pivot_wider(id_cols = c(gene_id, contrast),
              names_from = tissue,
              values_from = padj) %>%
  mutate(lrt_both = ifelse(!is.na(lrt) & !is.na(lrt12), 0, NA)) %>%
  pivot_longer(cols = c(lrt, lrt12, lrt_both),
               names_to = "tissue", values_to = "padj") %>%
  drop_na(padj) %>% 
  mutate(tissue = factor(tissue, levels = c("lrt", "lrt12", "lrt_both")))
```

```{r}
DE_lrt %>% 
  ggplot(aes(x = contrast, fill = tissue)) +
  geom_bar(position = "dodge", color = "grey40") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  scale_x_discrete(labels = contrast_labs) +
  scale_fill_brewer(palette = "Set1",
                    labels = c("3%", "12%", "both"),
                    name = "Sugar water") +
  labs(x = "", y = "Number of DEGs") +
  theme_classic() +
  theme(legend.position = "top")
```

### Venn diagrams of DEG overlap {.tabset .tabset-fade .tabset-pills}

```{r}
list_sugar <- function(fcontrast, DE_df) {
  sug03 <- DE_df %>%
    filter(sig == TRUE, contrast == fcontrast, tissue == "lrt") %>%
    pull(gene_id) %>%
    unique()
  sug12 <- DE_df %>%
    filter(sig == TRUE, contrast == fcontrast, tissue == "lrt12") %>%
    pull(gene_id) %>%
    unique()
  
  genes_list <- list(sug03, sug12)
  names(genes_list) <- c("3%", "12%") 
  return(genes_list)
}
```

```{r}
venn_sugar <- function(fcontrast, DE_df) {
  genes_list <- list_sugar(fcontrast, DE_df)
  if (all(lengths(genes_list) > 0)) {
    p <- venn.diagram(x = genes_list,
                      fill = contrast_cols,
                      cat.cex = 2, cat.fontface = "bold",
                      cat.pos = 0, cat.dist = 0.025,
                      main.cex = 1.6, cex = 2,
                      filename = NULL, disable.logging = TRUE)
    grid::grid.draw(p)
  }
}
```

<br>

#### LVP vs Virgin -- 6 hpm

```{r, results="hide", out.width="60%"}
venn_sugar(contrasts[1], DE_res_all)
```

<br>

#### LVP vs Virgin -- 24 hpm

```{r, results="hide", out.width="60%"}
venn_sugar(contrasts[2], DE_res_all)
```

<br>

### DEGs at both 3% and 12%  {.tabset .tabset-fade .tabset-pills}

<br>

```{r}
sugarcomp <- function(fcontrast, DE_df) {
  DE_df <- DE_df %>%
    select(-contrast_full, -trt_a, -trt_b, -sugar, -pvalue)
  
  sug03 <- DE_df %>%
    filter(sig == TRUE, contrast == fcontrast, tissue == "lrt") %>%
    select(-contrast, -description, -sig, -tissue)
  sug12 <- DE_df %>%
    filter(sig == TRUE, contrast == fcontrast, tissue == "lrt12") %>%
    select(-sig, -tissue)
  
  genes_df <- inner_join(sug03, sug12, by = "gene_id",
                         suffix = c("_03%", "_12%")) %>%
    select(contrast, gene_id, `lfc_03%`, `lfc_12%`,
           `padj_03%`, `padj_12%`, description) %>%
    mutate(lfc_mean = mean(c(`lfc_03%`, `lfc_12%`), na.rm = TRUE)) %>%
    arrange(lfc_mean) %>%
    select(-lfc_mean)
  
  return(genes_df)
}
```

```{r}
## Table with all genes significant in 12% or 3%
sugar_df <- map_dfr(.x = contrasts, .f = sugarcomp,
                    DE_res_all %>% filter(!grepl("akh", contrast)))
```

#### Table

```{r}
sugar_df %>%
  make_dt(numr_cols = c("lfc_03%", "padj_03%", "lfc_12%", "padj_12%"))
```

#### Heatmap for top-100

```{r}
if (nrow(sugar_df) > 1)
    plot_heat2(sugar_df$gene_id[1:min(100, length(sugar_df$gene_id))])
```

#### Boxplots for top-10

```{r}
walk(sugar_df$gene_id[1:min(10, length(sugar_df$gene_id))], plot_box)
```

### DEGs at 3% only

```{r}
DE_sel <- DE_res_all %>% select(-contrast_full, -trt_a, -trt_b, -sugar, -pvalue)
sug03_6hpm <- DE_sel %>% filter(sig == TRUE, contrast == "lvp06_vir06", tissue == "lrt")
sug12_6hpm <- DE_sel %>% filter(sig == TRUE, contrast == "lvp06_vir06", tissue == "lrt12")
sug03_24hpm <- DE_sel %>% filter(sig == TRUE, contrast == "lvp24_vir24", tissue == "lrt")
sug12_24hpm <- DE_sel %>% filter(sig == TRUE, contrast == "lvp24_vir24", tissue == "lrt12")
```

```{r}
sug03_only <- bind_rows(anti_join(sug03_6hpm, sug12_6hpm, by = "gene_id"),
                        anti_join(sug03_24hpm, sug12_24hpm, by = "gene_id"))
sug03_only %>% select(-sig, -tissue) %>% make_dt(numr_cols = c("lfc", "padj"))
```

### DEGs at 12% only

```{r}
sug12_only <- bind_rows(anti_join(sug12_6hpm, sug03_6hpm, by = "gene_id"),
                        anti_join(sug12_24hpm, sug03_24hpm, by = "gene_id"))
sug12_only %>% select(-sig, -tissue) %>% make_dt(numr_cols = c("lfc", "padj"))
```

### LFC correlations

Each point shows, for a single gene, the log-fold change values for a pairwise
contrast both in the at 3% and 12% sugar water.

```{r, out.width="90%"}
p <- sugar_df %>%
  ggplot(aes(x = `lfc_03%`, y = `lfc_12%`, color = contrast,
             text = glue("Gene: {gene_id}
                         Description: {description}
                         Contrast: {contrast}"))) +
  geom_point() +
  geom_hline(yintercept = 0, color = "grey20") +
  geom_vline(xintercept = 0, color = "grey20") +
  scale_color_brewer(palette = "Dark2", labels = contrast_labs) +
  theme_bw(base_size = 14) +
  labs(x = "Log-fold change - at 3%",
       y = "Log-fold change - at 12%",
       color = "timepoint")

p
#ggplotly(p, tooltip = "text")
```

```{r}
# Check the correlation at 6 hpm
sug6 <- sugar_df |> filter(contrast == "lvp06_vir06")

ggdensity(sug6$`lfc_03%`)
ggqqplot(sug6$`lfc_03%`)
shapiro.test(sug6$`lfc_03%`)
#W = 0.96551, p-value = 0.1432

ggdensity(sug6$`lfc_12%`)
ggqqplot(sug6$`lfc_12%`)
shapiro.test(sug6$`lfc_12%`)
#W = 0.97557, p-value = 0.3709

cor.test(x = sug6$`lfc_03%`, y = sug6$`lfc_12%`, method = "spearman")
#Spearman's rank correlation rho
#data:  sug6$`lfc_03%` and sug6$`lfc_12%`
#S = 3150, p-value < 2.2e-16
#alternative hypothesis: true rho is not equal to 0
#sample estimates:
#      rho 
#0.8574661 
```

```{r}
# Check the correlation at 24 hpm
sug24 <- sugar_df |> filter(contrast == "lvp24_vir24")

ggdensity(sug24$`lfc_03%`)
ggqqplot(sug24$`lfc_03%`)
shapiro.test(sug24$`lfc_03%`)
#W = 0.9044, p-value = 0.001954

ggdensity(sug24$`lfc_12%`)
ggqqplot(sug24$`lfc_12%`)
shapiro.test(sug24$`lfc_12%`)
#W = 0.84411, p-value = 4.476e-05

cor.test(x = sug24$`lfc_03%`, y = sug24$`lfc_12%`, method = "spearman")
#Spearman's rank correlation rho
#data:  sug24$`lfc_03%` and sug24$`lfc_12%`
#S = 376, p-value < 2.2e-16
#alternative hypothesis: true rho is not equal to 0
#sample estimates:
#      rho 
#0.9695325
```
