---
title: "Gene Ontology enrichment after aggregating DEGs by time point"
author: Jelmer Poelstra
date: today
format:
  html:
    cache: false
    embed-resources: true
    number-sections: true
    toc: true
    df-print: paged
    fig-dpi: 300
knitr:
  opts_chunk:
    message: FALSE
    warning: FALSE
    echo: FALSE
    out.width: "100%"
editor: visual
editor_options: 
  chunk_output_type: console
---

```{r}
# Load packages
library(tidyverse)
library(clusterProfiler)
library(enrichplot)
library(ggpubr)
library(here)
library(DT)

# Source scripts with helper functions
source(here("mcic-scripts/rnaseq/rfuns/enrich_funs.R"))
source(here("scripts/report_funs.R"))

# Define input files
go_res_file <- here("results/GO/GO_agglomByTime.tsv")

# Read input files
go_res <- read_tsv(go_res_file, show_col_types = FALSE) |>
  mutate(sig = ifelse(padj < 0.05 & numDEInCat > 1, TRUE, FALSE),
         fold_enrich = NA,
         mean_lfc = NA,
         median_lfc = NA) |>
  rename(n_DE_in_cat = numDEInCat,
         n_DE = numDE)

contrasts <- unique(go_res$contrast)
```

<br>

## Introduction

These are the GO enrichment results after aggregating DE genes across three tissues
(ab, ht, lrt) by timepoint.

For the purpose of this analysis,
any gene that is significantly DE in one or more of the four tissues
for each timepoint is counted as a DEG,
and genes that are not significant in any of the four tissues are counted as a
non-DEG.

After that aggregating step, the enrichment analysis was done the same way
as previously.

As previously, the enrichment test was done 3 times for each 'contrast'
(here, for each of the two timepoints):

- With all DEGs (`DE_direction` = "**both**")
- With DEGs more highly expressed in LVP than VIR (lfc > 0, "**up**")
- With DEGs more highly expressed in VIR than LVP (lfc < 0, "**down**")

<br>

## Summary table

```{r}
go_res |>
  count(contrast, DE_direction) |>
  left_join(go_res |> distinct(contrast, DE_direction, n_DE),
            by = c("contrast", "DE_direction")) |>
  rename(time = contrast, nr_sig_terms = n, nr_DEG = n_DE)
```

<br>

## Table with all significant terms

Some of the columns in the table below are:

- `padj` => Multiple-testing corrected p-value
- `nr_DEG_in_cat` => Number of DEGs in the focal GO category
- `nr_DEG` => Total number of DEGs for the focal contrast (timepoint)
- `nr_in_cat` => Total number of (expressed) genes in the GO category

```{r}
go_res |>
  select(time = contrast, DE_direction,
         category, ontology, description,
         padj, nr_DEG_in_cat = n_DE_in_cat,
         nr_in_cat = numInCat, nr_DEG = n_DE) |>
make_dt()
```

<br>

## Heatmap-style plots

### Up- and downregulated DEGs separately

- BP (Biological Processes) and CC (Cellular Components):

```{r}
go_res |>
  filter(DE_direction %in% c("up", "down"),
         ontology %in% c("BP", "CC")) |> 
  enrichplot(facet_var1 = "ontology",
             x_var = "DE_direction", facet_var2 = "contrast",
             ylab_size = 8)
```

- MF (Molecular Function):

```{r}
go_res |>
  filter(DE_direction %in% c("up", "down"),
         ontology %in% c("MF")) |> 
  enrichplot(facet_var1 = "ontology",
             x_var = "DE_direction", facet_var2 = "contrast",
             ylab_size = 8)
```

### All DEGs (either DE direction)

```{r}
go_res |>
  filter(DE_direction == "both") |>
  enrichplot(facet_var1 = "ontology",
             ylab_size = 8, countlab_size = 1.5)
```

## Cleveland's dotplots

### Up- and downregulated DEGs separately

- BP:

```{r}
go_res |>
  filter(DE_direction %in% c("up", "down"),
         ontology %in% c("BP")) |>
  cdotplot(facet_var1 = "contrast", facet_var2 = "DE_direction",
           fill_var = "padj_log", ylab_size = 7)
```

- CC:

```{r}
go_res |>
  filter(DE_direction %in% c("up", "down"),
         ontology %in% c("CC")) |>
  cdotplot(facet_var1 = "contrast", facet_var2 = "DE_direction",
           fill_var = "padj_log", ylab_size = 7)
```


- MF:

```{r}
go_res |>
  filter(DE_direction %in% c("up", "down"),
         ontology %in% c("MF")) |> 
  cdotplot(facet_var1 = "ontology", facet_var2 = "DE_direction",
           fill_var = "padj_log", ylab_size = 7)
```

### All DEGs (either DE direction)

```{r}
go_res |>
  filter(DE_direction == "both") |>
  cdotplot(facet_var1 = "ontology", facet_var2 = "contrast",
           fill_var = "padj_log", ylab_size = 7, point_size = 4.5) +
  scale_y_continuous(limits = c(0, NA),
                     expand = expansion(mult = c(0.005, 0.05)))
```
