---
title: |
  | _Aedes aegypti_ mRNA expression --
  | KEGG pathway analyses
pagetitle: "KEGG results"
author: "Jelmer Poelstra (poelstra.1@osu.edu), MCIC Wooster, OSU"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: cerulean
    highlight: kate
    toc: true
    toc_float: true
    fig_caption: true
    anchor_sections: true
    df_print: kable
    css: html_page.css
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  eval = TRUE,
  cache = FALSE,
  warning = FALSE,
  message = FALSE,
  out.width = "80%"
)
knitr::opts_knit$set(root.dir = here::here())
```

```{r}
## Install/load packages
if(! "pacman" %in% installed.packages()) install.packages("pacman")
packages <- c("tidyverse",        # Misc. data manipulation and plotting
              "here",             # Managing file paths
              "VennDiagram",      # Venn Diagram
              "ggpubr",           # Cleveland dot plots
              "ggforce",          # facet_col for equal tile heights
              "RColorBrewer",     # Plotting colors
              "plotly",           # Interactive plots
              "glue",             # Pasting text
              "DT", "kableExtra") # Tables             
pacman::p_load(char = packages)
```

```{r}
## Scripts with functions
source(here("scripts/GO_fun.R"))
source(here("scripts/kegg_fun.R"))
source(here("scripts/report_funs.R"))
```

```{r}
## Set contrasts of interest
contrasts <- c("lvp06_vir06", "lvp24_vir24")
contrast_labs <- c("6 hpm", "24 hpm")
contrasts_full <- c("lvp06_vir06_ab", "lvp24_vir24_ab",
                    "lvp06_vir06_ht", "lvp24_vir24_ht",
                    "lvp06_vir06_lrt", "lvp24_vir24_lrt",
                    "lvp06_vir06_lrt12", "lvp24_vir24_lrt12")
  
## Define tissues
tissues <- c("ab", "ht", "lrt", "lrt12")
```

```{r}
## Define input files
indir_DE <- here("results", "DE")
DE_file <- here(indir_DE, "DE_all.txt")
gene_file <- here(indir_DE, "gene_info.txt")

indir_kegg <- here("results", "kegg")
kegg_res_file <- here(indir_kegg, "kegg_all.txt")
kegg_map_file <- here(indir_kegg, "kegg_map.txt")
```

```{r}
## Read input files 1
gene_df <- read_tsv(gene_file, show_col_types = FALSE) %>%
  select(gene_id, description) %>%
  mutate(description = str_trunc(description, width = 50))

DE_res <- read_tsv(DE_file, show_col_types = FALSE) %>%
  mutate(tissue = gsub(".*\\d(\\w+)", "\\1", treat_a),
         tissue = ifelse(sugar == "sug12", "lrt12", tissue),
         trt_a = gsub("(.*\\d+)\\w+", "\\1", treat_a),
         trt_b = gsub("(.*\\d+)\\w+", "\\1", treat_b),
         contrast = paste0(trt_a, "_", trt_b),
         contrast_full = paste0(contrast, "_", tissue),
         sig = ifelse(padj < 0.05, TRUE, FALSE)) %>%
  filter(contrast %in% contrasts) %>%            # Only select focal contrasts!
  select(gene_id, lfc, pvalue, padj, sig,
         trt_a, trt_b, tissue, sugar, contrast, contrast_full) %>%
  left_join(gene_df, by = "gene_id")
```

```{r}
## Read input files 2
kegg_map <- read_tsv(kegg_map_file, show_col_types = FALSE) %>%
  rename(gene_id = geneID,
         category = KEGG_pathway)

kegg_res <- read_tsv(kegg_res_file, show_col_types = FALSE) %>%
  rename(numDEInCat = count) %>% 
  mutate(description = sub(" - Aedes.*", "", description),
         description = str_trunc(description, width = 50),
         description = str_wrap(description, width = 20),
         contrast = sub("lrt10", "lrt12", contrast),
         contrast_full = contrast,
         sig = ifelse(padj < 0.05, TRUE, FALSE),
         tissue = sub(".*\\d\\d_(\\w+)$", "\\1", contrast),
         contrast = sub("(.*_.*)_\\w+$", "\\1", contrast)) %>% 
  filter(contrast %in% contrasts)           # Only select focal contrasts!

kegg_lookup <- kegg_res %>%
  select(category, description) %>% 
  distinct()
```

```{r}
## Make a df with all pairwise contrasts
contrast_df <- DE_res %>%
  mutate(time_a = gsub(".*(\\d\\d)", "\\1", trt_a),
         time_b = gsub(".*(\\d\\d)", "\\1", trt_b),
         time_contrast = factor(paste0(time_a, "_", time_b),
                                levels = c("06_06", "24_24"))) %>%
  distinct(contrast_full, .keep_all = TRUE) %>% 
  select(contrast_full, contrast, trt_a, trt_b, tissue, sugar, time_contrast)
```

```{r, include=FALSE}
## Nr of genes with and without a KEGG annotation (added 2022-12-14)
length(unique(kegg_map$gene_id)) # 2,340
length(unique(DE_res$gene_id))   # 18,178

sig_genes <- DE_res %>% filter(sig == TRUE) %>% pull(gene_id) %>% unique()
length(sig_genes)                # 961
sum(sig_genes %in% unique(kegg_map$gene_id))  # 197
```

<br>

-----

## Methods

The R/Bioconductor package
[clusterProfiler 4.4.4](https://www.cell.com/the-innovation/fulltext/S2666-6758(21)00066-7)
was used to test for KEGG pathway enrichment among differentially expressed genes
for each pairwise contrast.
P-values were corrected for multiple testing using the Benjamini-Hochberg
correction.

<br>

-----

## Nr significant pathways per contrast

<br>

```{r}
sig_by_comp <- kegg_res %>%
  group_by(contrast, tissue, direction, .drop = FALSE) %>%
  count(name = "nsig") %>%
  ungroup() %>%
  pivot_wider(id_cols = c(contrast, direction),
              names_from = tissue, values_from = nsig,
              values_fill = 0) %>%
  left_join(distinct(contrast_df, contrast, .keep_all = TRUE),
            by = "contrast") %>% 
  arrange(time_contrast, trt_a, trt_b) %>%
  select(trt_a, trt_b, direction, ab, ht, lrt, lrt12)

sig_by_comp %>%
  rename("group A" = trt_a, "group B" = trt_b,
         "Ab" = ab, "HT" = ht, "LRT3" = lrt, "LRT12" = lrt12) %>%
  make_kable()
```

<br>

-----

## All significant pathways {.tabset .tabset-fade .tabset-pills}

<br>

### Both DE directions

<br>

```{r}
sig_smr <- pivot_wider(kegg_res %>% filter(direction == "both"),
                       id_cols = c(category, description),
                       names_from = contrast_full,
                       values_from = padj)

numr_cols <- colnames(sig_smr)[str_detect(colnames(sig_smr), "_")]
make_dt(sig_smr, numr_cols = numr_cols)
```

<br>

-----

### Upregulated after mating

<br>

```{r}
sig_smr <- pivot_wider(kegg_res %>% filter(direction == "up"),
                       id_cols = c(category, description),
                       names_from = contrast_full,
                       values_from = padj)

numr_cols <- colnames(sig_smr)[str_detect(colnames(sig_smr), "_")]
make_dt(sig_smr, numr_cols = numr_cols)
```

<br>

-----

### Downregulated after mating

<br>

```{r}
sig_smr <- pivot_wider(kegg_res %>% filter(direction == "down"),
                       id_cols = c(category, description),
                       names_from = contrast_full,
                       values_from = padj)

numr_cols <- colnames(sig_smr)[str_detect(colnames(sig_smr), "_")]
make_dt(sig_smr, numr_cols = numr_cols)
```

<br>

-----

## All DEGS in significant KEGG terms {.tabset .tabset-fade .tabset-pills}

<br>

```{r}
get_kegg_degs <- function(fcontrast, fdirection,
                          kegg_map, kegg_res, DE_res, gene_df) {
  
  message("Contrast: ", fcontrast, " / Direction: ", fdirection)
  
  fkegg_res <- kegg_res %>%
    filter(contrast_full == fcontrast,
           direction == fdirection) %>%
    select(-sig, -contrast_full) %>%
    rename(description_kegg = description)
  
  if (nrow(fkegg_res) == 0) return(NULL)
  
  kegg_res_wgenes <- fkegg_res %>%
    separate_wider_delim(gene_ids, delim = "/", names_sep = "_",
                         too_few = "align_start") %>% #, too_many = "debug"
    pivot_longer(cols = starts_with("gene_ids_"),
                 names_to = NULL, values_to = "gene_id",
                 values_drop_na = TRUE) %>%
    left_join(gene_df, by = "gene_id") %>%
    rename(description_gene = description) %>%
    relocate(tissue, .after = contrast) %>%
    relocate(gene_id, description_gene, .after = direction)
    
  #DEGs <- DE_res %>%
  #  filter(contrast_full == fcontrast, padj < 0.05) %>%
  #  pull(gene_id)
  
  #kegg_res_wgenes <- kegg_map %>%
  #  filter(gene_id %in% DEGs,
  #         category %in% fkegg_res$category) %>%
  #  full_join(fkegg_res, ., by = "category") %>% 
  #  left_join(gene_df, by = "gene_id") %>%
  #  rename(description_gene = description) %>%
  #  select(-gene_ids) %>% 
  #  relocate(tissue, .after = contrast) %>%
  #  relocate(gene_id, description_gene, .after = direction)
}
```

### Both DE directions

<br>

```{r}
kegg_degs <- map_dfr(.x = contrasts_full, .f = get_kegg_degs,
                     fdirection = "both",
                     kegg_map, kegg_res, DE_res, gene_df)
make_dt(kegg_degs)
```

<br>

-----

### Upregulated after mating

<br>

```{r}
kegg_degs <- map_dfr(.x = contrasts_full, .f = get_kegg_degs,
                     fdirection = "up",
                     kegg_map, kegg_res, DE_res, gene_df)
make_dt(kegg_degs)
```

<br>

-----

### Downregulated after mating

```{r}
kegg_degs <- map_dfr(.x = contrasts_full, .f = get_kegg_degs,
                     fdirection = "down",
                     kegg_map, kegg_res, DE_res, gene_df)
make_dt(kegg_degs)
```

<br>

-----

## Heatmaps - both DE directions {.tabset .tabset-fade .tabset-pills}

<br>

- **For DEGs in either direction (over- or underexpressed after mating).**

- Gray squares indicate non-significant values.
- Numbers indicate the number of DEGs in the focal category.
- If a timepopint is missing, there were no significant categories.

<br>

### All

```{r, fig.height = 6, out.width="100%"}
df <- prep_goplot(kegg_res, tissues = tissues, contrasts = contrasts) %>%
  mutate(contrast = sub("_ht.*|_lrt.*|_ab.*", "", contrast),
         tissue = sub("lrt12", "LRT 12%", tissue),
         tissue = sub("lrt", "LRT 3%", tissue),
         tissue = sub("ht", "HT", tissue),
         tissue = sub("ab", "Ab", tissue),
         tissue = factor(tissue,
                         levels = c("Ab", "HT", "LRT 3%", "LRT 12%"))) %>%
  filter(!is.na(padj)) %>%
  select(-padj, -category, -direction, -sig)

df %>%
  expand(description, tissue, contrast) %>%
  left_join(.,
            df %>% select(description) %>% distinct(),
            by = "description") %>% 
  left_join(df, by = c("description", "tissue", "contrast")) %>%
  goplot(type = "kegg", facet_var = "tissue",
         ylabsize = 10, numDE_size = 3.5) +
  scale_x_discrete(labels = contrast_labs)
```

<br>

-----

### Abdomen

```{r, fig.height = 6, out.width="100%"}
prep_goplot(kegg_res, tissues = "ab", contrasts) %>%
  goplot(type = "kegg", numDE_size = 3.5) +
  scale_x_discrete(labels = contrast_labs)
```

<br>

-----

### Head & thorax

```{r, out.width="100%"}
prep_goplot(kegg_res, tissues = "ht", contrasts) %>%
  goplot(type = "kegg", numDE_size = 3.5) +
  scale_x_discrete(labels = contrast_labs)
```

<br>

-----

### LRT (3%)

```{r, fig.height = 6, out.width="100%"}
prep_goplot(kegg_res, tissues = "lrt", contrasts) %>%
  goplot(type = "kegg", numDE_size = 3.5) +
  scale_x_discrete(labels = contrast_labs)
```

<br>

-----

### LRT 12%

```{r, fig.height = 6, out.width="100%"}
prep_goplot(kegg_res, tissues = "lrt12", contrasts) %>% 
  goplot(type = "kegg", numDE_size = 3.5) +
  scale_x_discrete(labels = contrast_labs)
```

<br>

-----

### LRT 3% + 12%

```{r, fig.height = 6, out.width="100%"}
df <- prep_goplot(kegg_res, tissues = c("lrt", "lrt12"),
                  contrasts = contrasts) %>%
  mutate(contrast = sub("_lrt.*|_ab.*", "", contrast),
         tissue = sub("lrt12", "LRT 12%", tissue),
         tissue = sub("lrt", "LRT 3%", tissue),
         tissue = factor(tissue, levels = c("LRT 3%", "LRT 12%"))) %>%
  filter(!is.na(padj)) %>%
  select(-padj, -category, -direction, -sig)

df %>%
  expand(description, tissue, contrast) %>%
  left_join(.,
            df %>% select(description) %>% distinct(),
            by = "description") %>% 
  left_join(df, by = c("description", "tissue", "contrast")) %>%
  goplot(type = "kegg", facet_var = "tissue",
         ylabsize = 8.5, numDE_size = 3.5) +
  scale_x_discrete(labels = contrast_labs)
```

<br>

-----

## Heatmaps - separate DE directions {.tabset .tabset-fade .tabset-pills}

<br>

**These plots have separate entries for enriched GO categories for down- and
upregulated genes after mating.**

- Gray squares indicate non-significant values.
- Numbers indicate the number of DEGs in the focal category.
- If a timepopint is missing, there were no significant categories.

<br>

### Abdomen

If omitted the DEG count labels for this plot because they were too squished
due to the high number of GO categories.

```{r, fig.height = 6, out.width="100%"}
prep_goplot(kegg_res, tissues = "ab", contrasts = contrasts,
            directions = c("up", "down")) %>%
  complete(category, contrast, direction) %>% 
  select(-description) %>%
  left_join(kegg_lookup, by = "category") %>% 
  mutate(contrast = ifelse(contrast == "lvp06_vir06", "06 hpm", "24 hpm")) %>% 
  goplot(type = "kegg", x_var = "direction", facet_var = "contrast",
         numDE_size = 3.5)
```

### Head & thorax

```{r, fig.height = 6, out.width="100%"}
prep_goplot(kegg_res, tissues = "ht", contrasts = contrasts,
            directions = c("up", "down")) %>%
  complete(category, contrast, direction) %>% 
  select(-description) %>%
  left_join(kegg_lookup, by = "category") %>% 
  mutate(contrast = ifelse(contrast == "lvp06_vir06", "06 hpm", "24 hpm")) %>% 
  goplot(type = "kegg", x_var = "direction", facet_var = "contrast",
         numDE_size = 3.5)
```

### LRT 3%

```{r, fig.height = 6, out.width="100%"}
prep_goplot(kegg_res, tissues = "lrt", contrasts = contrasts,
            directions = c("up", "down")) %>%
  complete(category, contrast, direction) %>% 
  select(-description) %>%
  left_join(kegg_lookup, by = "category") %>% 
  mutate(contrast = ifelse(contrast == "lvp06_vir06", "06 hpm", "24 hpm")) %>% 
  goplot(type = "kegg", x_var = "direction", facet_var = "contrast",
         numDE_size = 3.5)
```

### LRT 12%

```{r, fig.height = 6, out.width="100%"}
prep_goplot(kegg_res, tissues = "lrt12", contrasts = contrasts,
            directions = c("up", "down")) %>%
  complete(category, contrast, direction) %>% 
  select(-description) %>%
  left_join(kegg_lookup, by = "category") %>% 
  mutate(contrast = ifelse(contrast == "lvp06_vir06", "06 hpm", "24 hpm")) %>% 
  goplot(type = "kegg", x_var = "direction", facet_var = "contrast",
         numDE_size = 3.5)
```

<br>
